### 2021-05-03 paper/paper.tex
Outline for paper

\documentclass{article}

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[
backend=biber
]{biblatex}
\addbibresource{bibliography.bib}

\title{Algebraic Graphs with Class}
\author{
  Christoph Madlener\\
  \texttt{\href{mailto:madlener@in.tum.de}{madlener@in.tum.de}}
}

\begin{document}

\maketitle
\begin{abstract}
  ~\cite{mokhov2017algebraic} 
\end{abstract}

\section{Introduction}
\begin{itemize}
\item graph representations (in maths, in Haskell/other programming languages)
  \begin{itemize}
  \item ``standard'' tuple $G=(V,E), E \subseteq V \times V$
  \item containers~\cite{king1995dfs}, fgl~\cite{erwig2001inductive} (state of the art Haskell libraries)
  \end{itemize}
\item wellformedness (\textrightarrow{} implicit/explicit vertex set (?),
  partial functions)
\item introduce datatype (?)
\item sound and complete representation for graphs
\item overview
\end{itemize}

\section{Type class/locale}
\subsection{The Core}
\begin{itemize}
\item graph construction primitives
\item abstract from datatype \textrightarrow{} type class
\item fundamental functions (vertex, edge, etc.)
\end{itemize}
\subsection{Algebraic Structure}
\begin{itemize}
\item axioms (for digraphs)
\item subgraph - partial ordering
\item axioms for undirected graphs, other classes
\item instances in Haskell (directed, undirected, etc. via Eq instance)
\end{itemize}
\subsection{Graph Transformation Library}
\begin{itemize}
\item graph families (path, circuit, etc.)
\item transpose
\item functor
\item monad
\item removing edges
\end{itemize}
\subsection{Verification}
\begin{itemize}
\item equational reasoning
\item reasoning in Isabelle/HOL \textrightarrow{} locale
  \begin{itemize}
  \item ``advanced'' graph concepts (walks, reachability, SCCs, etc.)
    \begin{itemize}
    \item more experimentation needed for actual feasibility (additional axioms required
      for e.g.\ $Vertex\ u \neq \epsilon$?)
    \item first impression: similar as with other representations; abstraction
      to more advanced concepts quickly hides representation
    \end{itemize}
  \item ``polymorphic'' lemmas
  \item compare to Kruskal AFP entry~\cite{Kruskal-AFP} (possibility to get algorithms for
    any graph representation which can instantiate locale)
  \item instantiation with pair\_digraph~\cite{GraphTheory-AFP} - soundness and
    completeness proof
  \end{itemize}
\end{itemize}

\section{Deep Embedding}
\begin{itemize}
\item compact representation
\item compare to different concrete/executable representations wrt.\ efficiency
  in time and space 
  \begin{itemize}
  \item sparse graphs (edge list)
  \item adjacency map
  \end{itemize}
\item performance (\href{https://github.com/haskell-perf/graphs}{haskell-perf})
\item reasoning in Isabelle/HOL \textrightarrow{} induction is great
\item algorithms directly on deep embedding? \textrightarrow{} future work/open
  question (Haskell implementation converts to adjacency map for algorithms)
\end{itemize}
\subsection{Quotient Type}
\begin{itemize}
\item deep embedding which satisfies axioms
\item additional lemmas
\item minimization~\cite{mcconnell2005linear}
  \begin{itemize}
  \item interesting for quotient type, disconnect from Noschinski/any other
    representation for equality, same applies in Haskell for Eq instance)
    \item interesting for algorithms (assuming we can exploit compact
      representation)
    \end{itemize}
\end{itemize}

\section{Conclusion}
\begin{itemize}
\item summarize findings
\item restate open questions, future work
\item labeled graphs (outlook)
\item related work: other algebraic approaches (semiring on
  matrices~\cite{dolan2013fun}, relational
  algebra~\cite{berghammer2020relational})
\item usage examples~\cite{mokhov2019language,beaumont2017high}
\end{itemize}
\printbibliography
\end{document}

### 2021-05-04 paper/paper.tex
Draft of introduction

@@
 \usepackage{mathtools}
+\usepackage{minted}
+\setminted[haskell]{escapeinside=@@}
+\newcommand{\hs}{\mintinline{haskell}}
 \usepackage[hidelinks]{hyperref}
@@
 \begin{abstract}
-  ~\cite{mokhov2017algebraic} 
+  \cite{mokhov2017algebraic} 
 \end{abstract}
 
-\section{Introduction}
-\begin{itemize}
-\item graph representations (in maths, in Haskell/other programming languages)
-  \begin{itemize}
-  \item ``standard'' tuple $G=(V,E), E \subseteq V \times V$
-  \item containers~\cite{king1995dfs}, fgl~\cite{erwig2001inductive} (state of the art Haskell libraries)
-  \end{itemize}
-\item wellformedness (\textrightarrow{} implicit/explicit vertex set (?),
-  partial functions)
-\item introduce datatype (?)
-\item sound and complete representation for graphs
-\item overview
-\end{itemize}
+\section{Introduction}\label{sec:intro}
+Graphs are a fundamental structure studied in depth by both mathematicians and
+computer scientists alike. One very common definition states that a (directed)
+graph is a tuple $G = (V,E)$, where $V$ is a set of vertices and $E \subseteq V
+\times V$ is the set of edges. While this is a perfectly valid and natural
+mathematical definition it is not necessarily suitable for implementation. This
+can be illustrated by trying to directly translate this to Haskell:
+\begin{minted}{haskell}
+  data G a = G { vertices :: Set a, edges :: Set (a,a)}
+\end{minted}
+The value \hs{G {[1,2,3], [(1,2),(2,3)]}} then represents the graph $G =
+(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G {[1,2], [(2,3)]}} does not
+represent a consistent graph, as the edge refers to a non-existent node.
+The state-of-the-art \texttt{containers} library implements graphs with
+adjacency lists employing immutable arrays~\cite{king1995dfs}. The
+consistency condition $E \subseteq V \times V$ is not checked statically though,
+which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
+library uses inductive graphs~\cite{erwig2001inductive}, also exhibiting partial
+functions and potential for runtime errors due to the violation of consistency.
 
-\section{Type class/locale}
-\subsection{The Core}
+This lead Mokhov to conceive \textit{algebraic
+  graphs}~\cite{mokhov2017algebraic}, a sound and complete representation for
+graphs. They abstract away from graph representation details and characterize
+graphs by a set of axioms. Algebraic graphs have a small safe core of graph
+construction primitives and are suitable for implementing graph transformations.
+These primitives are represented in the following datatype:
+\begin{minted}{haskell}
+  data Graph a = Empty
+               | Vertex a
+               | Overlay (Graph a) (Graph a)
+               | Connect (Graph a) (Graph a)
+\end{minted}
+\hs{Empty} constructs the empty graph, \hs{Vertex} a graph with a single vertex
+and no edges. \hs{Overlay} essentially is the union of two graphs and
+\hs{Connect} additionally adds edges from all vertices from one graph to all
+vertices of the other.
+
+Alongside proper definitions for these primitives we will see
+in~\autoref{sec:locale} that this is indeed a sound and complete graph
+representation. We will also cover the algebraic structure (\ref{sec:algebra})
+exhibited by these graphs and elegant graph transformations (\ref{sec:trafo})
+based on a type class for the core. In~\autoref{sec:verification} we explore how
+we can exploit the algebraic structure in verification using Isabelle/HOL.
+
+In~\autoref{sec:deep} we will examine the deep embedding (i.e.\
+the datatype as opposed to the type class of the preceding sections) regarding
+applicability in practice and in verification. In~\autoref{sec:quotient} we
+define a quotient type in Isabelle/HOL which in essence is a deep embedding
+which satisfies the axioms.
+
+\section{Type class/locale}\label{sec:locale}
+\subsection{The Core}\label{sec:core}
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Algebraic Structure}
+\subsection{Algebraic Structure}\label{sec:algebra}
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Graph Transformation Library}
+\subsection{Graph Transformation Library}\label{sec:trafo}
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Verification}
+\subsection{Verification}\label{sec:verification}
 \begin{itemize}
@@
 
-\section{Deep Embedding}
+\section{Deep Embedding}\label{sec:deep}
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Quotient Type}
+\subsection{Quotient Type}\label{sec:quotient}
 \begin{itemize}
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 \begin{itemize}

### 2021-05-07 paper/paper.tex
Finish draft of type class section, start deep embedding section

@@
 
+\usepackage{geometry}
+% \geometry{left=3cm,right=3cm,top=3cm,bottom=3cm}
+
 \usepackage{graphicx}
+\usepackage{subcaption}
+
 \usepackage{mathtools}
+\usepackage{amssymb}
+
 \usepackage{minted}
@@
 \newcommand{\hs}{\mintinline{haskell}}
+
 \usepackage[hidelinks]{hyperref}
 \usepackage[utf8]{inputenc}
+\usepackage{enumitem}
+
 \usepackage[
@@
 \addbibresource{bibliography.bib}
+
+\renewcommand{\epsilon}{\varepsilon}
+\newcommand{\eps}{\epsilon}
+\newcommand{\overlay}{\oplus}
+\newcommand{\connect}{\rightarrow}
+
+\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
+\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
+
+% Swap the definition of \abs* and \norm*, so that \abs
+% and \norm resizes the size of the brackets, and the 
+% starred version does not.
+\makeatletter
+\let\oldabs\abs
+\def\abs{\@ifstar{\oldabs}{\oldabs*}}
+% 
+\let\oldnorm\norm
+\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
+\makeatother
 
@@
 
-\section{Type class/locale}\label{sec:locale}
+In~\autoref{sec:conclusion} we conclude and give an outlook on possible future
+work and open questions connected to algebraic graphs as presented.
+
+\section{A type class for algebraic graphs}\label{sec:locale}
+This section will cover most parts of the original paper by
+Mokhov~\cite{mokhov2017algebraic}. The graph construction primitives of
+algebraic graphs will be formally defined and their soundness and completeness
+proven. Afterwards we will cover the algebraic structure inherent to these
+primitives. As part of this we will introduce a type class in Haskell which
+further abstracts from implementation details. We will also see that different
+classes of graphs can be obtained by extending the set of axioms. Instances of
+these different classes and a graph transformation library in Haskell are
+provided. To conclude this section we will explore the suitability of this type
+class for formal verification in Isabelle/HOL.
+
 \subsection{The Core}\label{sec:core}
+In the introduction we already briefly saw the four graph construction
+primitives. Let us first define the set of (consistent) graphs $\mathcal{G}$
+over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
+then be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
+\subseteq V \times V$.
+The four graph construction primitives
+then are defined as follows. The \textit{empty} graph, denoted $\eps$ is
+simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
+$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
+\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
+clearly consistent, hence $v \in \mathcal{G}$. The binary operations
+\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
+respectively, allow constructing larger graphs. Their definitions are
+\begin{align*}
+  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
+  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
+\end{align*}
+So overlay is just the union of two graphs, while connect also adds an edge from
+each vertex of the ``left'' graph to each vertex of the ``right'' graph. It is
+straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
+then also $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect G_2 \in
+\mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
+graphs this already proves that algebraic graphs are a consistent
+representation for graphs. The completeness of this representation, i.e.\ for
+each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
+is shown in the next subsection. Both of these facts were also formally proven in
+Isabelle/HOL as part of~\autoref{sec:verification}.
+
+As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
+related to addition and multiplication, hence, we give connect a higher precedence
+than overlay. Let us consider some examples to develop some intuition for the
+primitives. They are illustrated in~\autoref{fig:construction}.
+\begin{enumerate}[label=(\alph*)]
+\item $1 \overlay 2 = (\{1,2\}, \emptyset)$
+\item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
+\item $1 \connect 2 \overlay 3 = (\{1,2,3\}, \{(1,2), (1,3)\})$
+\item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
+\end{enumerate}
+\begin{figure*}
+  \centering
+  \begin{subfigure}[b]{0.3\linewidth}
+    \centerline{\includegraphics[scale=0.27]{fig/ex-a.pdf}}
+    \vspace{-2.4mm}
+    \caption{$1 \overlay 2$}
+    \centerline{\includegraphics[scale=0.27]{fig/ex-b.pdf}}
+    \vspace{-2.4mm}
+    \caption{$1 \connect 2$}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\linewidth}
+    \centerline{\includegraphics[scale=0.27]{fig/ex-c.pdf}}
+    \vspace{-1mm}
+    \caption{$1 \connect (2 \overlay 3)$}
+  \end{subfigure}
+  \hspace{3mm}
+  \begin{subfigure}[b]{0.3\linewidth}
+    \centerline{\includegraphics[scale=0.27]{fig/ex-e-new.pdf}}
+    \vspace{-1mm}
+    \caption{$1 \connect 2 \overlay 2 \connect 3$}
+  \end{subfigure}
+  \caption{Examples of graph construction. The overlay and connect operations
+    are denoted by $\overlay$ and $\connect$, respectively. Illustrations
+    from~\cite{mokhov2017algebraic}.\label{fig:construction}} 
+\end{figure*}
+
+We already saw how to deeply embed these graph construction primitives in a
+datatype. For greater reusability the following type class can be defined.
+\begin{minted}{haskell}
+  class Graph g where
+      type Vertex g
+      empty :: g
+      vertex :: Vertex g -> g
+      overlay :: g -> g -> g
+      connect :: g -> g -> g
+\end{minted}
+The associated type \hs{Vertex g} represents the universe of vertices
+$\mathbb{V}$, the remainder of the type class corresponds to $\eps$, single
+vertex graphs, $\overlay$ resp.\ $\connect$.
+
+At this point we will only introduce some very basic functions for constructing
+graphs. More advanced constructions and transformations will be given
+in~\autoref{sec:trafo}. A graph with a single edge is obtained by simply
+connecting two vertices:
+\begin{minted}{haskell}
+  edge :: Graph g => Vertex g -> Vertex g -> g
+  edge u v = (vertex u) `connect` (vertex v)
+\end{minted}
+A graph with only isolated vertices can be constructed from a list of vertices
+as follows:
+\begin{minted}{haskell}
+  vertices :: Graph g => [Vertex g] -> g
+  vertices vs = foldr overlay empty . map vertex
+\end{minted}
+Replacing \hs{overlay} with \hs{connect} in \hs{vertices} leads to a (directed)
+clique. Later we will also consider undirected graphs. In that context this
+function actually builds a fully connected graph on the given list of vertices.
+\begin{minted}{haskell}
+  clique :: Graph g => [Vertex g] -> g
+  clique vs = foldr connect empty . map vertex
+\end{minted}
+Following and extending this approach leads to a safe and fully polymorphic
+graph transformation library for any concrete graph representation (e.g.\ graphs
+in \texttt{containers} or \texttt{fgl}) instantiating this class.
+
+\subsection{Algebraic Structure}\label{sec:algebra}
+Before turning to more advanced graph constructions we will observe the
+algebraic structure of the primitives. It turns out that overlay and connect
+almost form a semiring over graphs:
 \begin{itemize}
-\item graph construction primitives
-\item abstract from datatype \textrightarrow{} type class
-\item fundamental functions (vertex, edge, etc.)
+\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent commutative monoid
+\item $(\mathcal{G}, \connect, \eps)$ is a monoid
+\item $\connect$ distributes over $\overlay$
 \end{itemize}
-\subsection{Algebraic Structure}\label{sec:algebra}
+The shared identity and the missing annihilating zero for $\connect$
+(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
+\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
+\[
+  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
+\]
+Using decomposition we can actually prove some of the above properties, leading
+to the following minimal set of axioms characterizing directed graphs:
 \begin{itemize}
-\item axioms (for digraphs)
-\item subgraph - partial ordering
-\item axioms for undirected graphs, other classes
-\item instances in Haskell (directed, undirected, etc. via Eq instance)
+\item $(G, \overlay)$ is a commutative semigroup
+\item $(G, \connect, \eps)$ is a monoid
+\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
+  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
+  \connect z \overlay y \connect z$
+\item decomposition: $x \connect y \connect z = x \connect y \overlay x
+  \connect z \overlay y \connect z$ 
 \end{itemize}
+\begin{figure*}
+  \begin{subfigure}[b]{0.4\linewidth}
+    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
+    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
+      1 \connect 3$ }
+  \end{subfigure}
+  \hspace{12mm}
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
+    \caption{Decomposition: $1 \connect 2 \connect 3 = {1 \connect 2} \overlay
+      1 \connect 3 \overlay 2 \connect 3$}
+  \end{subfigure}
+  \vspace{-1mm}
+  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
+\end{figure*}
+One can easily check that the definitions of the primitives given
+in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
+completeness algebraic graphs. For any graph $G=(V,E)$ the following function
+builds it using the four construction primitives:
+\begin{minted}{haskell}
+  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
+  graph vs es = overlay (vertices vs) (edges es)
+\end{minted}
+The \hs{edges} function generalizes \hs{edge} to a list of edges:
+\begin{minted}{haskell}
+  edges :: Graph g => [(Vertex g, Vertex g)] -> g
+  edges = foldr overlay empty . map (uncurry edge)
+\end{minted}
+This was also formally proven in Isabelle/HOL (see~\autoref{sec:verification}).
+
+Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
+idempotent overlay operation as $x \preceq y \iff x \overlay y = y$. This is indeed
+a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
+transitive). In fact this partial order can actually be used as a definition for
+the subgraph relation:
+\[
+  x \subseteq y \coloneqq x \preceq y \iff x \overlay y = y
+\]
+
+So far we only considered directed graphs. Undirected graphs can be
+obtained from the same construction primitives as directed graphs. It suffices
+to modify the underlying axioms. In the case of undirected graphs this can be
+achieved by making connect commutative (then denoted by $\leftrightarrow$),
+i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
+can prove some of the axioms of directed graphs. So undirected graphs are
+characterized by the following minimal set of axioms:
+\begin{itemize}
+\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
+\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
+\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
+  y \oplus x \leftrightarrow z$
+\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
+  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
+\end{itemize}
+By choosing different additional axioms, other classes of graphs can be
+characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
+self-loop), transitive graphs (as in dependency graphs), and the combination
+thereof; preorders. Combining undirected graphs and reflexive graphs yields
+equivalence relations. It is even possible to define hypergraphs of arbitrary
+(but fixed) order by replacing the decomposition law with another suitable
+axiom~\cite{mokhov2017algebraic}.
+
+In summary the algebraic approach to graph representation has a small and safe
+core of construction primitives. Pairing these primitives with suitable sets of
+axioms leads to a highly flexible representation for graphs.
+
 \subsection{Graph Transformation Library}\label{sec:trafo}
-\begin{itemize}
-\item graph families (path, circuit, etc.)
-\item transpose
-\item functor
-\item monad
-\item removing edges
-\end{itemize}
+Based on this core type class, Mokhov implements what he refers to as a graph
+transformation library\footnote{Algebraic graphs on hackage:
+  \texttt{\href{http://hackage.haskell.org/package/algebraic-graphs}{http://hackage.haskell.org/package/algebraic-graphs}}}.
+In order to understand the following constructions and transformations better,
+we follow the approach of Mokhov and first define instances for the type class
+\hs{Graph}. Recall the attempt from the introduction to directly translate the
+graph representation $G=(V,E)$ into the Haskell datatype \hs{G}. We will now
+define a \hs{Graph} instance for that datatype, while realizing that more
+generally it can be used to represent relations on some domain (hence the renaming
+to \hs{Relation}).
+\begin{minted}{haskell}
+data Relation a = R { domain :: Set a, relation :: Set (a, a)} deriving Eq
+
+instance Ord a => Graph (Relation a) where
+  type Vertex (Relation a) = a
+  empty       = R Set.empty Set.empty
+  vertex  x   = R (singleton x) Set.empty
+  overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
+  connect x y = R (domain x `union` domain y) (relation x `union` relation y `union'
+    fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
+\end{minted}
+Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
+of \hs{+} and \hs{*} as shortcuts for \hs{overlay} resp.\ \hs{connect}.
+
+The given \hs{Graph} instance for \hs{Relation} satisfies the axioms of directed
+graphs, as was argued in~\autoref{sec:core}. Obtaining \hs{Graph} instances
+which satisfy the axioms for undirected graphs (or other classes of graphs) is
+also straightforward: One can simply wrap \hs{Relation} in a \hs{newtype} and
+define a custom \hs{Eq} instance. An example for undirected graphs (in this case
+think of symmetric directed graphs) is given.
+\begin{minted}{haskell}
+newtype Symmetric a = S (Relation a) deriving (Graph, Num)
+
+instance Ord a => Eq (Symmetric a) where
+  S x == S y = symmetricClosure x == symmetricClosure y
+\end{minted}
+It is also sensible to introduce subclasses like \hs{class Graph g => UndirectedGraph g}
+to reflect the class of graph we are working with in the type system, and thus,
+increase type safety.
+
+Equipped with these two instances let us now actually consider graph
+constructions and transformations made possible with the core type class. Simple
+examples include constructing a \textit{path} from a list of vertices, which
+then can be easily extended to a \textit{circuit}:
+\begin{minted}{haskell}
+path :: Graph g => [Vertex g] -> g
+path []  = empty
+path [x] = vertex x
+path xs = edges $ zip xs (tail xs)
+%$
+circuit :: Graph g => [Vertex g] -> g
+circuit [] = empty
+circuit xs = path (xs ++ [head xs])
+\end{minted}
+Mokhov also gives functions for constructing \textit{star} graphs (i.e.\ a
+center vertex connected to a set of leaves) and for building graphs from
+\hs{Tree}s and \hs{Forest}s from the \texttt{containers} library.
+
+More interesting applications include the implementation of a zero time
+graph transposition using a \hs{newtype} wrapper~\cite{mokhov2017algebraic}:
+\begin{minted}{haskell}
+newtype Transpose g = T { transpose :: g } deriving Eq
+
+instance Graph g => Graph (Transpose g) where
+  type Vertex (Transpose g) = Vertex g
+  empty       = T empty
+  vertex      = T . vertex
+  overlay x y = T $ overlay (transpose x) (transpose y)
+  connect x y = T $ connect (transpose y) (transpose x)
+\end{minted}
+Although for certain representations it is possible to get an $\mathcal{O}(1)$
+transpose operation (e.g.\ by setting a transposed flag), for many typical
+representations the whole graph has to be traversed, resulting in
+$\mathcal{O}(\abs{V} + \abs{E})$ time.
+
+More sophisticated transformations can be implemented using the following
+\hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
+\hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
+vertices of type \hs{b} polymorphically.
+\begin{minted}{haskell}
+newtype GraphFunctor a = F { gfor :: forall g. Graph g => (a -> Vertex g) -> g }
+
+instance Graph (GraphFunctor a) where
+  type Vertex (GraphFunctor a) = a
+  empty       = F $ \_ -> empty
+  vertex  x   = F $ \f -> vertex (f x)
+  overlay x y = F $ \f -> overlay (gmap f x) (gmap f y)
+  connect x y = F $ \f -> connect (gmap f x) (gmap f y)
+
+gmap :: Graph g => (a -> Vertex g) -> GraphFunctor a -> g
+gmap = flip gfor  
+\end{minted}
+Interestingly this can be used to merge vertices fulfilling some predicate
+\hs{p} by mapping them to a single vertex:
+\begin{minted}{haskell}
+mergeVertices :: Graph g => (Vertex g -> Bool)
+  -> Vertex g -> GraphFunctor (Vertex g) -> g
+mergeVertices p v = gmap $ \u -> if p u then v else u
+\end{minted}
+In the original paper \hs{gmap} is then used further to construct rather
+sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}.
+
+Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
+\hs{bind} in the context of graphs allows to replace each vertex with a
+(possibly) empty subgraph. This can be used to both remove and split vertices.
+\begin{minted}{haskell}
+newtype GraphMonad a = M { bind :: forall g. Graph g => (a -> g) -> g }
+
+instance Graph (GraphMonad a) where
+  type Vertex (GraphMonad a) = a
+  empty       = M $ \_ -> empty
+  vertex x    = M $ \f -> f x
+  overlay x y = M $ \f -> overlay (bind x f) (bind y f)
+  connect x y = M $ \f -> connect (bind x f) (bind y f)
+\end{minted}
+Removing a vertex with \hs{bind} can be generalized to a function \hs{induce},
+which only keeps vertices satisfying some predicate \hs{p}.
+\begin{minted}{haskell}
+induce :: Graph g => (Vertex g -> Bool) -> GraphMonad (Vertex g) -> g
+induce p g = bind g $ \v -> if p v then vertex v else empty
+\end{minted}
+Splitting a vertex works by replacing that vertex with a graph containing the
+desired vertices.
+\begin{minted}{haskell}
+splitVertex :: (Graph g, Eq (Vertex g)) => Vertex g -> [Vertex g] 
+  -> GraphMonad (Vertex g) -> g
+splitVertex v vs g = bind g $ \u -> if u == v then vertices vs else vertex u
+\end{minted}
+Mokhov then defines yet another \hs{newtype} wrapper utilizing \hs{bind} to
+remove edges. He also showcases the construction capabilities of the presented
+library by constructing de Bruijn graphs (a rather sophisticated type of graph
+occurring frequently occurring in computer engineering and bioinformatics)
+completely polymorphically. These definitions are omitted here for brevity.
+
+The presented material on the given type class gives an overview of an elegant,
+reusable and polymorphic graph construction and transformation library. Note
+that in the available \texttt{hackage} library for algebraic graphs the
+\hs{newtype} wrappers for graph transformation are not implemented in the
+presented type class. Instead there is a higher-kinded type class which is based
+on \hs{MonadPlus}, but has fewer instances~\cite{mokhov2017algebraic}.
+
 \subsection{Verification}\label{sec:verification}
-\begin{itemize}
-\item equational reasoning
-\item reasoning in Isabelle/HOL \textrightarrow{} locale
-  \begin{itemize}
-  \item ``advanced'' graph concepts (walks, reachability, SCCs, etc.)
-    \begin{itemize}
-    \item more experimentation needed for actual feasibility (additional axioms required
-      for e.g.\ $Vertex\ u \neq \epsilon$?)
-    \item first impression: similar as with other representations; abstraction
-      to more advanced concepts quickly hides representation
-    \end{itemize}
-  \item ``polymorphic'' lemmas
-  \item compare to Kruskal AFP entry~\cite{Kruskal-AFP} (possibility to get algorithms for
-    any graph representation which can instantiate locale)
-  \item instantiation with pair\_digraph~\cite{GraphTheory-AFP} - soundness and
-    completeness proof
-  \end{itemize}
-\end{itemize}
+We have now seen how to implement a polymorphic graph transformation library in
+Haskell using the core type class. Mokhov proposes that algebraic graphs and
+their axiomatic characterization are suitable for formal verification as they
+enable equational reasoning. Some
+simple statements (like $\hs{vertices xs} \subseteq \hs{clique xs}$) are given
+accompanied by an Agda formalization\footnote{available at
+  \texttt{\href{https://github.com/snowleopard/alga-theory}{https://github.com/snowleopard/alga-theory}}}.
+For the examples in that formalization the claim obviously holds.
+
+To experiment further we first formalized the available material in
+Isabelle/HOL~\cite{isabelle}.
+This formalization\footnote{available at
+  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
+involves the definition of a \textit{locale}, which for the
+purposes of this paper can be thought of as a type class which can be equipped
+with axioms. Proving the statements from the paper in Isabelle/HOL works
+conveniently.
+
+Noschinski formalized graph theory in
+Isabelle/HOL~\cite{GraphTheory-AFP}. A part of that is essentially the standard
+representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with a
+consistency axiom). Instantiating a locale (which can be likened to
+instantiating a type class in Haskell) requires to prove the axioms attached to
+it. Using the definitions from~\autoref{sec:core} for the primitives it is
+straightforward to instantiate the algebraic graph locale for the tuple
+representation. We also formally prove that the graph construction primitives only
+produce consistent graphs. A proof for the completeness of the primitives is
+given as well (as long as we accept that the tuple representation is a
+\textit{complete} graph representation).
+
+As the statements in the original paper were of very limited scope, we 
+define and reason about more \textit{advanced} graph theoretic concepts: We
+introduce (vertex) walks and the notion of reachability on algebraic graphs.
+First impressions suggest that it is feasible to base more sophisticated
+formalizations of graph theory on algebraic graphs. Although we suspect that
+some additional axiom(s) may be required to make for \textit{interesting} graph
+theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
+\mathbb{V}$. This is further illustrated by the fact that we can instantiate the
+algebraic graph locale with the unit type.
+
+This is an interesting direction for future work for a related reason that the
+type class is interesting in Haskell: The type class is the basis for 
+polymorphic graph transformation library, whereas the locale can be
+seen as a \textit{polymorphic} graph formalization library, as any lemma we can prove
+using the locale axioms become available to a graph representation
+instantiating it. This obviously becomes more interesting the more material is
+available in the locale. Haslbeck et al.\ chose a similar approach to implement
+and verify Kruskal's algorithm independent of the concrete graph
+representation~\cite{Kruskal-AFP}.
 
 \section{Deep Embedding}\label{sec:deep}
+Up until this point we were mainly concerned with exploiting the algebraic
+structure of graphs for polymorphic graph constructions and transformations, and
+for formal verification. It turns out, however, that the deep embedding of the
+construction primitives in a datatype is interesting in itself. Recall the
+definition from the introduction:
+\begin{minted}{haskell}
+data Graph a = Empty
+             | Vertex a
+             | Overlay (Graph a) (Graph a)
+             | Connect (Graph a) (Graph a)
+\end{minted}
+An instance for the type class can be defined easily, however, we cannot use the
+derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
+One solution is to fold the datatype into a concrete graph representation to
+obtain a law-abiding \hs{Eq} instance:
+\begin{minted}{haskell}
+fold :: Graph g => Graph (Vertex g) -> g
+fold Empty         = empty
+fold (Vertex  x  ) = vertex x
+fold (Overlay x y) = overlay (fold x) (fold y)
+fold (Connect x y) = connect (fold x) (fold y)
+
+instance Ord a => Eq (Graph a) where
+  x == y = fold x == (fold y :: Relation a)
+\end{minted}
+This approach is also currently implemented in the library available on
+\texttt{hackage}. However, this is not ideal for multiple reasons. Coupling the
+datatype to some other graph representation merely for equality checks is
+inconvenient at best. More interestingly algebraic graphs can compactly
+represent dense graphs. Consider the graph \hs{clique [1..n]}, which is linear
+in size in the deep embedding, while it is quadratic when stored as a
+\hs{Relation}. Requiring to fully expand graphs to check equality voids this
+advantage.
 \begin{itemize}

### 2021-05-08 paper/paper.tex
Finish first draft of paper

@@
 \begin{abstract}
-  \cite{mokhov2017algebraic} 
+  We give an overview of algebraic graphs as introduced by
+  Mokhov~\cite{mokhov2017algebraic}. Algebraic graphs have a small and safe core
+  of construction primitives, alongside a set of axioms, characterizing an
+  algebra of graphs. They can be used to elegantly implement a graph
+  transformation library employing functional programming. We also review the
+  suitability of algebraic graphs for formal verification using Isabelle/HOL.
 \end{abstract}
@@
 in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
-completeness algebraic graphs. For any graph $G=(V,E)$ the following function
+completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
 builds it using the four construction primitives:
@@
 transitive). In fact this partial order can actually be used as a definition for
-the subgraph relation:
+the subgraph relation, denoted by $\subseteq$:
 \[
@@
 i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
-can prove some of the axioms of directed graphs. So undirected graphs are
+can prove some of the axioms of directed graphs. Hence, undirected graphs are
 characterized by the following minimal set of axioms:
@@
   path xs = edges $ zip xs (tail xs)
-%$
+
   circuit :: Graph g => [Vertex g] -> g
@@
 \hs{bind} in the context of graphs allows to replace each vertex with a
-(possibly) empty subgraph. This can be used to both remove and split vertices.
+(possibly empty) subgraph. This can be used to both remove and split vertices.
 \begin{minted}{haskell}
@@
 theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
-\mathbb{V}$. This is further illustrated by the fact that we can instantiate the
-algebraic graph locale with the unit type.
+\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
+instantiate the algebraic graph locale with the unit type.
 
 This is an interesting direction for future work for a related reason that the
-type class is interesting in Haskell: The type class is the basis for 
-polymorphic graph transformation library, whereas the locale can be
-seen as a \textit{polymorphic} graph formalization library, as any lemma we can prove
-using the locale axioms become available to a graph representation
+type class is interesting in Haskell: The type class forms the basis for 
+a polymorphic graph transformation library, whereas the locale can be
+seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
+using the locale axioms becomes available to any graph representation
 instantiating it. This obviously becomes more interesting the more material is
@@
 structure of graphs for polymorphic graph constructions and transformations, and
-for formal verification. It turns out, however, that the deep embedding of the
+for formal verification. However, that the deep embedding of the
 construction primitives in a datatype is interesting in itself. Recall the
@@
 derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
-One solution is to fold the datatype into a concrete graph representation to
+One solution is to fold the deep embedding into a concrete graph representation to
 obtain a law-abiding \hs{Eq} instance:
@@
 \hs{Relation}. Requiring to fully expand graphs to check equality voids this
-advantage.
-\begin{itemize}
-\item compact representation
-\item compare to different concrete/executable representations wrt.\ efficiency
-  in time and space 
-  \begin{itemize}
-  \item sparse graphs (edge list)
-  \item adjacency map
-  \end{itemize}
-\item performance (\href{https://github.com/haskell-perf/graphs}{haskell-perf})
-\item reasoning in Isabelle/HOL \textrightarrow{} induction is great
-\item algorithms directly on deep embedding? \textrightarrow{} future work/open
-  question (Haskell implementation converts to adjacency map for algorithms)
-\end{itemize}
-\subsection{Quotient Type}\label{sec:quotient}
-\begin{itemize}
-\item deep embedding which satisfies axioms
-\item additional lemmas
-\item minimization~\cite{mcconnell2005linear}
-  \begin{itemize}
-  \item interesting for quotient type, disconnect from Noschinski/any other
-    representation for equality, same applies in Haskell for Eq instance)
-    \item interesting for algorithms (assuming we can exploit compact
-      representation)
-    \end{itemize}
-\end{itemize}
+advantage. A promising approach to alleviate this issue is the \textit{modular
+  graph decomposition}~\cite{mcconnell2005linear}, as it allows to minimize
+algebraic graph expressions. It remains an open question if this decomposition
+can be computed efficiently directly on the deep embedding.
+
+More interesting questions for future work come up in connection with the deep
+embedding. Currently only the transformations presented with the type class, i.e.\
+transpose, as well as anything that can be done using the corresponding versions
+of \hs{gmap} and \hs{bind}, are implemented directly on the deep embedding. No
+known efficient implementations of fundamental graph algorithms, like
+depth-first-search, that work directly on the deep embedding are
+known~\cite{mokhov2017algebraic}. The library therefore resorts to converting
+algebraic graph expressions to an adjacency map and uses algorithms
+for those.
+
+An immediate application which can be more efficient in space using
+algebraic graphs that comes to mind is search, however, only in best case
+examples, i.e.\ where only a small part of the graph has to be expanded. This is
+due to the fact that the neighborhood of a vertex in an algebraic graph
+expression can be computed in linear (in the size of the graph expression) time
+and space. We already saw that the algebraic graph representation can be
+significantly smaller than conventional representations.
+
+Some benchmarks were performed (on an earlier version of the library) to compare
+its performance to \texttt{containers} and \texttt{fgl} (and one other
+representation). The results are mixed; depending on the area of application,
+algebraic graphs may be competitive in performance, or in some cases, orders of
+magnitudes slower. For the full results refer to the benchmark suite
+repository\footnote{available at
+  \texttt{\href{https://github.com/haskell-perf/graphs}{https://github.com/haskell-perf/graphs}}}.
+\subsection{Verification}\label{sec:quotient}
+We were also interested in using the deep embedding in formal verification using
+Isabelle/HOL. Hence, we formalized most of the material from the library. This
+feat turned out nicely, as the inductive datatype definition of the deep
+embedding enables proofs by induction.
+
+In the Haskell implementation of the deep embedding we already saw that
+structural equality does not satisfy the axioms. There we can define a custom
+\hs{Eq} instance to obtain a law-abiding datatype. In Isabelle/HOL we can follow
+a similar approach, which results in a \texttt{quotient\_type}, which can be
+thought of as an equivalence class construction on the raw datatype. For the
+moment the equivalence is based on the tuple representation by Noschinski
+mentioned in~\autoref{sec:verification}. Again, the minimization of algebraic
+graph expressions for an equivalence relation on algebraic graphs, decoupled
+from any other graph representation, is desirable.
+
+The quotient type also allows us to formally prove that it satisfies the axioms,
+by instantiating the locale. It also paves the way for further formalization
+efforts on the deep embedding. However, this only becomes really interesting
+with more future work, i.e.\ if algorithms that work directly on the deep
+embedding or other advanced usages of it can be found (in other words, if there
+is more material to formalize).
 
 \section{Conclusion}\label{sec:conclusion}
-\begin{itemize}
-\item summarize findings
-\item restate open questions, future work
-\item labeled graphs (outlook)
-\item related work: other algebraic approaches (semiring on
-  matrices~\cite{dolan2013fun}, relational
-  algebra~\cite{berghammer2020relational})
-\item usage examples~\cite{mokhov2019language,beaumont2017high}
-\end{itemize}
+We saw a novel approach to representing graphs, which is especially suitable for
+functional programming. It allows for safe, polymorphic and elegant graph
+constructions and transformations. Algebraic graphs also lend themselves nicely
+to formal verification.
+
+The alert reader might have noticed that the given primitives can only be used
+to represent unlabeled graphs. Mokhov extended the algebraic graph library to
+also represent edge-labeled graphs. There is no published material on this yet
+(besides the source code), for more information refer to the talk by Mokhov
+during Haskell eXchange 2018\footnote{available at
+  \texttt{\href{https://skillsmatter.com/skillscasts/12361-labelled-algebraic-graphs}{https://skillsmatter.com/skillscasts/12361-labelled-algebraic-graphs}}}.
+
+Algebraic graphs as presented have been successfully used in practice, e.g.\ for
+asynchronous circuit design~\cite{beaumont2017high} and acceleration of graph
+processing using FPGAs~\cite{mokhov2019language}. These are examples for which
+the available transformations are sufficient. An important direction for future
+research is going beyond those (relatively simple) transformations. That means
+to find and formulate algorithms either inside the core type class or directly
+on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}
+successfully employ algebraic approaches to formally verify graph algorithms.
+However, they are using an algebra of relations as opposed to the one here
+defined using graph construction primitives. Dolan~\cite{dolan2013fun} uses an
+algebra on adjacency matrices to find paths in graphs.
+Mokhov himself postulates that the algebraic approach allows to formulate graph
+algorithms as systems of equations with unknowns. This approach may also be a
+way to find novel graph algorithms.
+
 \printbibliography
+
 \end{document}

### 2021-05-10 paper/paper.tex
Minor changes to paper

@@
 \end{minted}
-The value \hs{G {[1,2,3], [(1,2),(2,3)]}} then represents the graph $G =
-(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G {[1,2], [(2,3)]}} does not
+The value \hs{G [1,2,3] [(1,2),(2,3)]} then represents the graph $G =
+(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G [1,2] [(2,3)]} does not
 represent a consistent graph, as the edge refers to a non-existent node.
@@
 the datatype as opposed to the type class of the preceding sections) regarding
-applicability in practice and in verification. In~\autoref{sec:quotient} we
-define a quotient type in Isabelle/HOL which in essence is a deep embedding
-which satisfies the axioms.
+applicability in practice and in verification (\autoref{sec:quotient}).
 
@@
 algebraic graphs will be formally defined and their soundness and completeness
-proven. Afterwards we will cover the algebraic structure inherent to these
-primitives. As part of this we will introduce a type class in Haskell which
+proven. Afterwards we will cover the algebraic structure exhibited by these
+primitives when interpreted as graphs. As part of this we will introduce a type class in Haskell which
 further abstracts from implementation details. We will also see that different
@@
 over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
-then be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
+be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
 \subseteq V \times V$.
@@
 \mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
-clearly consistent, hence $v \in \mathcal{G}$. The binary operations
+trivially consistent, hence $v \in \mathcal{G}$. The binary operations
 \textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
@@
 So overlay is just the union of two graphs, while connect also adds an edge from
-each vertex of the ``left'' graph to each vertex of the ``right'' graph. It is
+each vertex of its first argument to each vertex of its second argument. It is
 straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
@@
 is shown in the next subsection. Both of these facts were also formally proven in
-Isabelle/HOL as part of~\autoref{sec:verification}.
+Isabelle/HOL (see~\autoref{sec:verification}).
 
@@
 \item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
-\item $1 \connect 2 \overlay 3 = (\{1,2,3\}, \{(1,2), (1,3)\})$
+\item $1 \connect (2 \overlay 3) = (\{1,2,3\}, \{(1,2), (1,3)\})$
 \item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
@@
 \begin{itemize}
-\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent commutative monoid
+\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
 \item $(\mathcal{G}, \connect, \eps)$ is a monoid
@@
     \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
-    \caption{Decomposition: $1 \connect 2 \connect 3 = {1 \connect 2} \overlay
-      1 \connect 3 \overlay 2 \connect 3$}
+    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
+      {1 \connect 3} \overlay {2 \connect 3}$}
   \end{subfigure}
@@
 \end{minted}
-This was also formally proven in Isabelle/HOL (see~\autoref{sec:verification}).
+They are a direct translation of the fact, that any algebraic graph expression
+$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
+\textit{canonical form}:
+\[
+  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
+  u \connect v \right)
+\]
+The completeness of algebraic graphs was also formally proven in Isabelle/HOL
+(see~\autoref{sec:verification}).
 
@@
 So far we only considered directed graphs. Undirected graphs can be
-obtained from the same construction primitives as directed graphs. It suffices
+obtained from the very same construction primitives as directed graphs. It suffices
 to modify the underlying axioms. In the case of undirected graphs this can be
 achieved by making connect commutative (then denoted by $\leftrightarrow$),
-i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
-can prove some of the axioms of directed graphs. Hence, undirected graphs are
-characterized by the following minimal set of axioms:
+i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
+introducing this axiom we can prove some of the other axioms of directed graphs.
+Hence, undirected graphs are characterized by the following minimal set of
+axioms:
 \begin{itemize}
@@
 \subsection{Graph Transformation Library}\label{sec:trafo}
-Based on this core type class, Mokhov implements what he refers to as a graph
+Based on the core type class, Mokhov implements what he refers to as a graph
 transformation library\footnote{Algebraic graphs on hackage:
@@
 In order to understand the following constructions and transformations better,
-we follow the approach of Mokhov and first define instances for the type class
+we follow his approach and first define instances for the type class
 \hs{Graph}. Recall the attempt from the introduction to directly translate the
@@
 define a custom \hs{Eq} instance. An example for undirected graphs (in this case
-think of symmetric directed graphs) is given.
+the connection to symmetric directed graphs is very prominent) is given.
 \begin{minted}{haskell}
@@
 More interesting applications include the implementation of a zero time
-graph transposition using a \hs{newtype} wrapper~\cite{mokhov2017algebraic}:
+graph transposition using yet another \hs{newtype}
+wrapper~\cite{mokhov2017algebraic}:
 \begin{minted}{haskell}
@@
 \hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
-vertices of type \hs{b} polymorphically.
+vertices of type \hs{b}.
 \begin{minted}{haskell}
@@
 Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
-\hs{bind} in the context of graphs allows to replace each vertex with a
+\hs{bind} function in the context of graphs allows to replace each vertex with a
 (possibly empty) subgraph. This can be used to both remove and split vertices.
@@
 library by constructing de Bruijn graphs (a rather sophisticated type of graph
-occurring frequently occurring in computer engineering and bioinformatics)
+occurring frequently in computer engineering and bioinformatics)
 completely polymorphically. These definitions are omitted here for brevity.
@@
 that in the available \texttt{hackage} library for algebraic graphs the
-\hs{newtype} wrappers for graph transformation are not implemented in the
-presented type class. Instead there is a higher-kinded type class which is based
+\hs{newtype} wrappers for graph transformation are not implemented as such.
+Instead there is a higher-kinded type class which is based
 on \hs{MonadPlus}, but has fewer instances~\cite{mokhov2017algebraic}.
@@
 with axioms. Proving the statements from the paper in Isabelle/HOL works
-conveniently.
+without problems.
 
-Noschinski formalized graph theory in
-Isabelle/HOL~\cite{GraphTheory-AFP}. A part of that is essentially the standard
-representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with a
-consistency axiom). Instantiating a locale (which can be likened to
+Noschinski formalized fundamental aspects of graph theory in
+Isabelle/HOL~\cite{GraphTheory-AFP}. Part of that is essentially the standard
+representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with one single
+axiom ensuring consistency). Instantiating a locale in Isabelle/HOL (which can be likened to
 instantiating a type class in Haskell) requires to prove the axioms attached to
@@
 produce consistent graphs. A proof for the completeness of the primitives is
-given as well (as long as we accept that the tuple representation is a
-\textit{complete} graph representation).
+given as well (in the sense that every graph represented by a tuple can be
+expressed as an algebraic graph).
 
@@
 some additional axiom(s) may be required to make for \textit{interesting} graph
-theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
+theory; as it stands the statement $\eps \neq \hs{vertex u}$ for some $u \in
 \mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
@@
 using the locale axioms becomes available to any graph representation
-instantiating it. This obviously becomes more interesting the more material is
+instantiating the locale. This obviously becomes more interesting the more material is
 available in the locale. Haslbeck et al.\ chose a similar approach to implement
@@
 structure of graphs for polymorphic graph constructions and transformations, and
-for formal verification. However, that the deep embedding of the
+for formal verification. However, the deep embedding of the
 construction primitives in a datatype is interesting in itself. Recall the
@@
 This approach is also currently implemented in the library available on
-\texttt{hackage}. However, this is not ideal for multiple reasons. Coupling the
-datatype to some other graph representation merely for equality checks is
-inconvenient at best. More interestingly algebraic graphs can compactly
-represent dense graphs. Consider the graph \hs{clique [1..n]}, which is linear
-in size in the deep embedding, while it is quadratic when stored as a
-\hs{Relation}. Requiring to fully expand graphs to check equality voids this
-advantage. A promising approach to alleviate this issue is the \textit{modular
-  graph decomposition}~\cite{mcconnell2005linear}, as it allows to minimize
-algebraic graph expressions. It remains an open question if this decomposition
-can be computed efficiently directly on the deep embedding.
+\texttt{hackage} (although an adjacency map is used instead of a relation).
+However, this is not ideal for multiple reasons. Coupling the datatype to some
+other graph representation merely for equality checks is inconvenient at best.
+More interestingly algebraic graphs can compactly represent dense graphs.
+Consider the graph \hs{clique [1..n]}, which has a linear size representation in
+the deep embedding, while it is quadratic when stored as a \hs{Relation}.
+Requiring to fully expand graphs to check equality voids this advantage. A
+promising approach to alleviate this issue is \textit{modular graph
+  decomposition}~\cite{mcconnell2005linear}, as it allows to minimize algebraic
+graph expressions. It remains an open question if this decomposition can be
+computed efficiently directly on the deep embedding.
 
@@
 of \hs{gmap} and \hs{bind}, are implemented directly on the deep embedding. No
-known efficient implementations of fundamental graph algorithms, like
-depth-first-search, that work directly on the deep embedding are
-known~\cite{mokhov2017algebraic}. The library therefore resorts to converting
-algebraic graph expressions to an adjacency map and uses algorithms
-for those.
+known efficient implementations of fundamental graph algorithms - like
+depth-first-search - that work directly on the deep embedding are
+known~\cite{mokhov2017algebraic}. The current library implementation therefore
+resorts to converting algebraic graph expressions to an adjacency map and reuses
+existing algorithms on this representation.
 
-An immediate application which can be more efficient in space using
-algebraic graphs that comes to mind is search, however, only in best case
+An immediate application which can be more efficient in space while also being
+competitive regarding time complexity using
+algebraic graphs that comes to mind is search. However, this only applies in best case
 examples, i.e.\ where only a small part of the graph has to be expanded. This is
@@
 
-Some benchmarks were performed (on an earlier version of the library) to compare
+Benchmarks (on an earlier version of the library) were performed by Moine to compare
 its performance to \texttt{containers} and \texttt{fgl} (and one other
 representation). The results are mixed; depending on the area of application,
-algebraic graphs may be competitive in performance, or in some cases, orders of
-magnitudes slower. For the full results refer to the benchmark suite
-repository\footnote{available at
+algebraic graphs offer a competitive alternative, while in other cases, they
+perform orders of magnitudes worse. For the full results refer to the repository
+containing the benchmark suite\footnote{available at
   \texttt{\href{https://github.com/haskell-perf/graphs}{https://github.com/haskell-perf/graphs}}}.
+
 \subsection{Verification}\label{sec:quotient}
 We were also interested in using the deep embedding in formal verification using
-Isabelle/HOL. Hence, we formalized most of the material from the library. This
-feat turned out nicely, as the inductive datatype definition of the deep
-embedding enables proofs by induction.
+Isabelle/HOL. Hence, we once more formalized large parts of the material on the
+deep embedding from the Haskell library. This
+feat turned out to be very convenient, as the inductive datatype definition of the deep
+embedding enables proofs by induction. Many of the resulting subgoals were then handled
+automatically by Isabelle.
 
@@
 thought of as an equivalence class construction on the raw datatype. For the
-moment the equivalence is based on the tuple representation by Noschinski
-mentioned in~\autoref{sec:verification}. Again, the minimization of algebraic
-graph expressions for an equivalence relation on algebraic graphs, decoupled
-from any other graph representation, is desirable.
+moment the equivalence is based on conversion of algebraic graph expressions to
+the tuple representation by Noschinski mentioned in~\autoref{sec:verification}.
+Again, the minimization of algebraic graph expressions to obtain an equivalence
+relation on algebraic graphs, decoupled from any other graph representation, is
+very much desirable.
 
 The quotient type also allows us to formally prove that it satisfies the axioms,
-by instantiating the locale. It also paves the way for further formalization
-efforts on the deep embedding. However, this only becomes really interesting
-with more future work, i.e.\ if algorithms that work directly on the deep
-embedding or other advanced usages of it can be found (in other words, if there
-is more material to formalize).
+by instantiating the locale corresponding to the core type class. It also paves
+the way for further formalization efforts on the deep embedding. However, this
+also requires more work to become truly interesting, i.e.\ if algorithms that
+work directly on the deep embedding or other advanced usages of it can be found
+(or put rather bluntly, if there is more material to formalize).
 
@@
 asynchronous circuit design~\cite{beaumont2017high} and acceleration of graph
-processing using FPGAs~\cite{mokhov2019language}. These are examples for which
+processing using FPGAs~\cite{mokhov2019language}. In these examples
 the available transformations are sufficient. An important direction for future
@@
 However, they are using an algebra of relations as opposed to the one here
-defined using graph construction primitives. Dolan~\cite{dolan2013fun} uses an
-algebra on adjacency matrices to find paths in graphs.
+defined on graph construction primitives. Dolan~\cite{dolan2013fun} uses an
+algebra on adjacency matrices (specifically a semiring) to find paths in graphs.
 Mokhov himself postulates that the algebraic approach allows to formulate graph
-algorithms as systems of equations with unknowns. This approach may also be a
-way to find novel graph algorithms.
+algorithms as systems of equations with unknowns. This may also be an approach
+for finding novel graph algorithms.
 

### 2021-05-27 paper/paper.tex
Incorporate feedback from reviews

@@
 
-\usepackage{geometry}
+\usepackage[a4paper]{geometry}
 % \geometry{left=3cm,right=3cm,top=3cm,bottom=3cm}
@@
 
+\usepackage{isabelle,isabellesym}
+
+\newcommand*{\term}[1]{{\isaspacing\isastyle \input{#1}}}
+\newcommand*{\Term}[1]{{\isaspacing\isastyle #1}}
+\newcommand*{\mTerm}[1]{\text{\Term{#1}}}
+
 \renewcommand{\epsilon}{\varepsilon}
@@
 \DeclarePairedDelimiter\norm{\lVert}{\rVert}%
+
+\let\subsectionautorefname\sectionautorefname
+\let\subsubsectionautorefname\sectionautorefname
 
@@
 
-This lead Mokhov to conceive \textit{algebraic
+This led Mokhov to conceive \textit{algebraic
   graphs}~\cite{mokhov2017algebraic}, a sound and complete representation for
@@
 construction primitives and are suitable for implementing graph transformations.
-These primitives are represented in the following datatype:
-\begin{minted}{haskell}
-  data Graph a = Empty
-               | Vertex a
-               | Overlay (Graph a) (Graph a)
-               | Connect (Graph a) (Graph a)
-\end{minted}
-\hs{Empty} constructs the empty graph, \hs{Vertex} a graph with a single vertex
-and no edges. \hs{Overlay} essentially is the union of two graphs and
-\hs{Connect} additionally adds edges from all vertices from one graph to all
-vertices of the other.
+The four construction primitives are the following:
+\begin{enumerate}
+\item the \emph{empty} graph
+\item single \emph{vertex} graphs
+\item \emph{overlay}ing two graphs
+\item \emph{connect}ing two graphs
+\end{enumerate}
+Overlay is the union of two graphs while connect also adds
+edges from all vertices from one graph to all vertices of the other.
 
@@
 
-In~\autoref{sec:deep} we will examine the deep embedding (i.e.\
-the datatype as opposed to the type class of the preceding sections) regarding
-applicability in practice and in verification (\autoref{sec:quotient}).
+In~\autoref{sec:deep} we will examine the \emph{deep embedding} of the construction
+primitives into an inductive datatype and its
+applicability in practice and in verification (\ref{sec:quotient}).
 
@@
 
-\section{A type class for algebraic graphs}\label{sec:locale}
+\section{A Type Class for Algebraic Graphs}\label{sec:locale}
 This section will cover most parts of the original paper by
@@
 further abstracts from implementation details. We will also see that different
-classes of graphs can be obtained by extending the set of axioms. Instances of
+classes of graphs can be obtained by modifying the axioms. Instances of
 these different classes and a graph transformation library in Haskell are
-provided. To conclude this section we will explore the suitability of this type
+presented. To conclude this section we will explore the suitability of this type
 class for formal verification in Isabelle/HOL.
@@
 \subsection{The Core}\label{sec:core}
-In the introduction we already briefly saw the four graph construction
-primitives. Let us first define the set of (consistent) graphs $\mathcal{G}$
-over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
-be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
-\subseteq V \times V$.
-The four graph construction primitives
-then are defined as follows. The \textit{empty} graph, denoted $\eps$ is
+Before defining the four graph construction primitives we define the set of
+graphs $\mathcal{G}$ over a fixed universe of vertices $\mathbb{V}$. A graph $G
+\in \mathcal{G}$ can be represented as a tuple $G=(V,E)$ where $V \subseteq
+\mathbb{V}$. A graph is \emph{consistent} if $E \subseteq V \times V$ holds.
+
+The construction primitives
+then are defined as follows. The \textit{empty} graph, denoted by $\eps$, is
 simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
 $\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
-\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
+\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$, denoted by just $v$, are also
 trivially consistent, hence $v \in \mathcal{G}$. The binary operations
@@
 straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
-then also $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect G_2 \in
-\mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
+are consistent, then so are $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect
+G_2 \in \mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
 graphs this already proves that algebraic graphs are a consistent
-representation for graphs. The completeness of this representation, i.e.\ for
+representation for graphs, i.e.\ using the construction primitives only yields
+consistent graphs. The completeness of this representation, i.e.\ for
 each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
-is shown in the next subsection. Both of these facts were also formally proven in
+is shown shortly (\ref{sec:algebra}). Both of these facts were also formally proven in
 Isabelle/HOL (see~\autoref{sec:verification}).
@@
 As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
-related to addition and multiplication, hence, we give connect a higher precedence
-than overlay. Let us consider some examples to develop some intuition for the
-primitives. They are illustrated in~\autoref{fig:construction}.
+related to addition and multiplication in a semiring, hence, we give connect a
+higher precedence
+than overlay. Let us consider a few examples to develop some intuition for the
+primitives. In these examples the universe of vertices are the natural numbers
+and they are illustrated in~\autoref{fig:construction}.
 \begin{enumerate}[label=(\alph*)]
@@
 
-We already saw how to deeply embed these graph construction primitives in a
-datatype. For greater reusability the following type class can be defined.
+We model the graph construction primitives in Haskell with the following type
+class:
 \begin{minted}{haskell}
@@
 $\mathbb{V}$, the remainder of the type class corresponds to $\eps$, single
-vertex graphs, $\overlay$ resp.\ $\connect$.
+vertex graphs, $\overlay$ and $\connect$ respectively.
 
@@
   vertices :: Graph g => [Vertex g] -> g
-  vertices vs = foldr overlay empty . map vertex
+  vertices = foldr overlay empty . map vertex
 \end{minted}
@@
   clique :: Graph g => [Vertex g] -> g
-  clique vs = foldr connect empty . map vertex
+  clique = foldr connect empty . map vertex
 \end{minted}
@@
 \end{itemize}
-The shared identity and the missing annihilating zero for $\connect$
-(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
+For a proper semiring an annihilating zero for $\connect$ (i.e.\
+$x \connect 0 = 0 = 0 \connect x$) is missing. It is also unusual that the two operations
+share their identity $\eps$.
+
+In addition there is the following
 \textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
@@
 \]
-Using decomposition we can actually prove some of the above properties, leading
-to the following minimal set of axioms characterizing directed graphs:
-\begin{itemize}
-\item $(G, \overlay)$ is a commutative semigroup
-\item $(G, \connect, \eps)$ is a monoid
-\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
-  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
-  \connect z \overlay y \connect z$
-\item decomposition: $x \connect y \connect z = x \connect y \overlay x
-  \connect z \overlay y \connect z$ 
-\end{itemize}
 \begin{figure*}
@@
 \end{figure*}
-One can easily check that the definitions of the primitives given
+With decomposition we can prove by equational reasoning that $\overlay$ is
+idempotent and that $\eps$ is indeed its identity.
+According to Mokhov, the remaining axioms are a minimal set of axioms
+\emph{characterizing} algebraic graphs~\cite{mokhov2017algebraic}. From the
+original paper it does not become clear what he means by characterization:
+Letting $\mathcal{G} = \{1\}$, $\eps = 1$, $1 \overlay 1 = 1$ and
+$1\connect 1 = 1$ gives a structure abiding all axioms. This definition of
+$\mathcal{G}$ is clearly not (isomorphic to) what we call directed graphs for
+any universe of vertices.
+We don't know which additional axioms are required to obtain this isomorphism.
+However, the statement $\eps \neq vertex\ u$ for any $u \in \mathbb{V}$ does not
+follow from the given axioms. Hence, adding an axiom scheme $\forall u \in
+\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ probably
+needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$), however, we don't know if this
+is sufficient.
+
+One can, however, easily verify that the definitions of the primitives given
 in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
@@
 \]
-The completeness of algebraic graphs was also formally proven in Isabelle/HOL
-(see~\autoref{sec:verification}).
 
 Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
-idempotent overlay operation as $x \preceq y \iff x \overlay y = y$. This is indeed
+idempotent overlay operation as $x \preceq y \vcentcolon\Leftrightarrow x \overlay y = y$. This is indeed
 a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
-transitive). In fact this partial order can actually be used as a definition for
+transitive). In fact this partial order can be used as a definition for
 the subgraph relation, denoted by $\subseteq$:
 \[
-  x \subseteq y \coloneqq x \preceq y \iff x \overlay y = y
+  x \subseteq y \vcentcolon\Leftrightarrow x \preceq y \vcentcolon\Leftrightarrow x \overlay y = y
 \]
@@
 introducing this axiom we can prove some of the other axioms of directed graphs.
-Hence, undirected graphs are characterized by the following minimal set of
-axioms:
+Hence, according to Mokhov~\cite{mokhov2017algebraic}, undirected graphs are
+characterized by the following minimal set of axioms:
 \begin{itemize}
@@
 characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
-self-loop), transitive graphs (as in dependency graphs), and the combination
+self-loop), transitive graphs (as used in dependency graphs), and the combination
 thereof; preorders. Combining undirected graphs and reflexive graphs yields
 equivalence relations. It is even possible to define hypergraphs of arbitrary
-(but fixed) order by replacing the decomposition law with another suitable
-axiom~\cite{mokhov2017algebraic}.
+(but fixed) rank by replacing the decomposition law with another suitable
+axiom~\cite{mokhov2017algebraic}. In all of these the same caveats regarding the
+term \emph{characterization} from directed graphs apply.
 
@@
     overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
-    connect x y = R (domain x `union` domain y) (relation x `union` relation y `union'
+    connect x y = R (domain x `union` domain y) (relation x `union` relation y `union`
       fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
@@
 In the original paper \hs{gmap} is then used further to construct rather
-sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}.
+sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}, i.e.\
+independent from any concrete graph representation.
 
@@
 
-To experiment further we first formalized the available material in
+To experiment further we first formalized the material from the paper in
 Isabelle/HOL~\cite{isabelle}.
@@
   \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
-involves the definition of a \textit{locale}, which for the
-purposes of this paper can be thought of as a type class which can be equipped
-with axioms. Proving the statements from the paper in Isabelle/HOL works
+the definition of several \textit{locale}s. For the purposes of this paper a
+locale be thought of as a type class which can be equipped with axioms. The
+following locale is the basis for our formalization.
+
+\vspace{3mm}
+\term{terms/locale_primitives}
+\vspace{3mm}\newline
+Note that it does not have any
+axioms, as we reuse it to define locales for different classes of graphs (i.e.\
+directed, undirected, etc.). Proving the statements from the paper in Isabelle/HOL works
 without problems.
@@
 given as well (in the sense that every graph represented by a tuple can be
-expressed as an algebraic graph).
+obtained using the algebraic graph construction primitives).
 
@@
 define and reason about more \textit{advanced} graph theoretic concepts: We
-introduce (vertex) walks and the notion of reachability on algebraic graphs.
-First impressions suggest that it is feasible to base more sophisticated
-formalizations of graph theory on algebraic graphs. Although we suspect that
-some additional axiom(s) may be required to make for \textit{interesting} graph
-theory; as it stands the statement $\eps \neq \hs{vertex u}$ for some $u \in
-\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
-instantiate the algebraic graph locale with the unit type.
+introduce (vertex) walks and the notion of reachability on algebraic graphs. A
+challenge posed by the approach is that a graph is simply a value of some
+abstract type \Term{'g}. Thus, we characterize the existence of a vertex or an
+edge in a given graph by the subgraph relation (denoted by $\sqsubseteq$ in the
+formalization):
 
-This is an interesting direction for future work for a related reason that the
-type class is interesting in Haskell: The type class forms the basis for 
-a polymorphic graph transformation library, whereas the locale can be
-seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
-using the locale axioms becomes available to any graph representation
-instantiating the locale. This obviously becomes more interesting the more material is
-available in the locale. Haslbeck et al.\ chose a similar approach to implement
-and verify Kruskal's algorithm independent of the concrete graph
-representation~\cite{Kruskal-AFP}.
+\vspace{3mm}
+\term{terms/has_elem}
+\vspace{3mm}\newline
+Equipped with these definitions we can then go on to inductively define when a
+list of vertices is a walk in a given graph:
+
+\vspace{3mm}
+\term{terms/vwalk_def}
+\vspace{3mm}\newline
+We define the empty list as well as the list containing only a single vertex of
+the graph as walks (\Term{vwalk\_Nil} and \Term{vwalk\_single}). Additionally if
+a list with head \Term{v} (note that Haskell's \hs{(:)} is denoted by \Term{(\#)} in
+Isabelle) is a walk and $(u,v) \in E$, then prepending \Term{u} also yields a
+walk. Now we can prove various statements about walks (e.g.\ appending walks,
+splitting walks, etc.). Interestingly, many of these statements follow directly
+from the inductive definition of walks, i.e.\ we don't require any of the
+algebraic graph axioms to prove them. We only need them when a proof relies on
+the fact that $(u,v) \in E$ implies $u \in V$ as well as $v \in V$ (in other
+words, the consistency of a graph).
+
+For reachability we took two approaches. First we defined the set of edges of a
+graph as follows:
+
+\vspace{3mm}
+\term{terms/edge_set_def}
+\vspace{3mm}\newline
+Reachability then can be defined via the reflexive, transitive closure of the
+edge set. The second approach is to say that $v$ is reachable from $u$ if and
+only if there is a walk from $u$ to $v$. These are, however, only proof of
+concepts as only transitivity of reachability has been shown for both definitions.
+
+Nonetheless, first impressions from these experiments suggest that it is
+feasible to base more sophisticated formalizations of graph theory on algebraic
+graphs. We did not experiment with adding additional axioms as described
+in~\autoref{sec:algebra}. We instantiate the locale with the unit type, where
+all primitives simply map to the unit value. This formalizes that the current
+set of axioms is insufficient.
+
+Exploring verification using the locale further is an interesting direction for
+future work. The Haskell type class forms the basis for a polymorphic graph
+transformation library. In the same vein, the locale can be seen as a
+\textit{polymorphic} graph formalization
+library. Any lemma we can prove using the locale axioms becomes available to any
+graph representation instantiating the locale. This obviously becomes more
+interesting the more material is available in the locale. Haslbeck et al.\ chose
+a similar approach to implement and verify Kruskal's algorithm independent of
+the concrete graph representation~\cite{Kruskal-AFP}.
 
@@
 An instance for the type class can be defined easily, however, we cannot use the
-derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
+derived \hs{Eq} instance for the datatype, as it would violate the graph axioms:
+\hs{Overlay Empty Empty} and \hs{Empty} both represent $\eps$ but are not deemed
+equal.
 One solution is to fold the deep embedding into a concrete graph representation to
@@
 However, this is not ideal for multiple reasons. Coupling the datatype to some
-other graph representation merely for equality checks is inconvenient at best.
+other graph representation merely for equality checks is undesirable.
+
 More interestingly algebraic graphs can compactly represent dense graphs.
@@
   decomposition}~\cite{mcconnell2005linear}, as it allows to minimize algebraic
-graph expressions. It remains an open question if this decomposition can be
-computed efficiently directly on the deep embedding.
+graph expressions.
 
@@
 of \hs{gmap} and \hs{bind}, are implemented directly on the deep embedding. No
-known efficient implementations of fundamental graph algorithms - like
+efficient implementations of fundamental graph algorithms - like
 depth-first-search - that work directly on the deep embedding are
@@
 
-An immediate application which can be more efficient in space while also being
-competitive regarding time complexity using
-algebraic graphs that comes to mind is search. However, this only applies in best case
-examples, i.e.\ where only a small part of the graph has to be expanded. This is
-due to the fact that the neighborhood of a vertex in an algebraic graph
-expression can be computed in linear (in the size of the graph expression) time
-and space. We already saw that the algebraic graph representation can be
-significantly smaller than conventional representations.
+One application that comes to mind is search. Consider a best case example,
+where only a small part of a dense graph has to be expanded before finding a target:
+This allows us to exploit the better space complexity (linear vs.\ quadratic)
+while also being competitive regarding time. The neighborhood of a vertex in an
+algebraic graph can be computed in linear (in the size of the graph expression)
+time and space.
 
 Benchmarks (on an earlier version of the library) were performed by Moine to compare
-its performance to \texttt{containers} and \texttt{fgl} (and one other
-representation). The results are mixed; depending on the area of application,
-algebraic graphs offer a competitive alternative, while in other cases, they
-perform orders of magnitudes worse. For the full results refer to the repository
-containing the benchmark suite\footnote{available at
+the performance of the deep embedding to \texttt{containers} and \texttt{fgl}
+(and one other representation). The results vary heavily depending on the area of
+application. In some cases algebraic graphs offer a competitive alternative,
+while in others, they perform orders of magnitudes worse. For the full results
+refer to the repository containing the benchmark suite\footnote{available at
   \texttt{\href{https://github.com/haskell-perf/graphs}{https://github.com/haskell-perf/graphs}}}.
@@
 We were also interested in using the deep embedding in formal verification using
-Isabelle/HOL. Hence, we once more formalized large parts of the material on the
-deep embedding from the Haskell library. This
-feat turned out to be very convenient, as the inductive datatype definition of the deep
+Isabelle/HOL. Hence, we ported the functions to Isabelle/HOL and then proved
+their correctness. This
+turned out to be very convenient, as the inductive datatype definition of the deep
 embedding enables proofs by induction. Many of the resulting subgoals were then handled
@@
 also requires more work to become truly interesting, i.e.\ if algorithms that
-work directly on the deep embedding or other advanced usages of it can be found
-(or put rather bluntly, if there is more material to formalize).
+work directly on the deep embedding or other advanced usages of it can be found.
 
@@
 also represent edge-labeled graphs. There is no published material on this yet
-(besides the source code), for more information refer to the talk by Mokhov
+(besides the source code). For more information refer to the talk by Mokhov
 during Haskell eXchange 2018\footnote{available at
@@
 to find and formulate algorithms either inside the core type class or directly
-on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}
+on the deep embedding. Berghammer et. al~\cite{berghammer2020relational}, as well
+as, Guttmann and Robinson-O'Brien~\cite{Relational_Minimum_Spanning_Trees-AFP}
 successfully employ algebraic approaches to formally verify graph algorithms.

### 2021-05-27 paper/terms/edge_set_def.tex
Incorporate feedback from reviews

\isacommand{definition}\isamarkupfalse%
\ edge{\isacharunderscore}{\kern0pt}set\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}v\ {\isasymtimes}\ {\isacharprime}{\kern0pt}v{\isacharparenright}{\kern0pt}\ set{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ {\isachardoublequoteopen}edge{\isacharunderscore}{\kern0pt}set\ g\ {\isasymequiv}\ {\isacharbraceleft}{\kern0pt}{\isacharparenleft}{\kern0pt}u{\isacharcomma}{\kern0pt}v{\isacharparenright}{\kern0pt}\ {\isacharbar}{\kern0pt}\ u\ v{\isachardot}{\kern0pt}\ has{\isacharunderscore}{\kern0pt}edge\ g\ u\ v{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}\isanewline

### 2021-05-27 paper/terms/has_elem.tex
Incorporate feedback from reviews

\isacommand{definition}\isamarkupfalse%
\ has{\isacharunderscore}{\kern0pt}vertex\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}v\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \isakeyword{where}\ \isanewline
\ \ {\isachardoublequoteopen}has{\isacharunderscore}{\kern0pt}vertex\ g\ u\ {\isasymequiv}\ vertex\ u\ {\isasymsqsubseteq}\ g{\isachardoublequoteclose}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\ has{\isacharunderscore}{\kern0pt}edge\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}v\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}v\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \isakeyword{where}\ \isanewline
\ \ {\isachardoublequoteopen}has{\isacharunderscore}{\kern0pt}edge\ g\ u\ v\ {\isasymequiv}\ edge\ u\ v\ {\isasymsqsubseteq}\ g{\isachardoublequoteclose}

### 2021-05-27 paper/terms/locale_primitives.tex
Incorporate feedback from reviews

\isacommand{locale}\isamarkupfalse%
\ algebraic{\isacharunderscore}{\kern0pt}pre{\isacharunderscore}{\kern0pt}graph\ {\isacharequal}{\kern0pt}\ \isanewline
\ \ \isakeyword{fixes}\ empty\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacharprime}{\kern0pt}g\ {\isacharparenleft}{\kern0pt}{\isachardoublequoteopen}{\isasymepsilon}{\isachardoublequoteclose}{\isacharparenright}{\kern0pt}\isanewline
\ \ \ \ \isakeyword{and}\ vertex\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}v\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\isanewline
\ \ \ \ \isakeyword{and}\ overlay\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\ {\isacharparenleft}{\kern0pt}\isakeyword{infixl}\ {\isacartoucheopen}{\isasymoplus}{\isacartoucheclose}\ {\isadigit{7}}{\isadigit{5}}{\isacharparenright}{\kern0pt}\isanewline
\ \ \ \ \isakeyword{and}\ connect\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\ {\isacharparenleft}{\kern0pt}\isakeyword{infixl}\ {\isacartoucheopen}{\isasymrightarrow}{\isacartoucheclose}\ {\isadigit{8}}{\isadigit{0}}{\isacharparenright}{\kern0pt}

### 2021-05-27 paper/terms/vwalk_def.tex
Incorporate feedback from reviews

\isacommand{inductive}\isamarkupfalse%
\ vwalk\ \isakeyword{where}\isanewline
\ \ vwalk{\isacharunderscore}{\kern0pt}Nil{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}vwalk\ g\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbar}{\kern0pt}\isanewline
\ \ vwalk{\isacharunderscore}{\kern0pt}single{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}has{\isacharunderscore}{\kern0pt}vertex\ g\ u\ {\isasymLongrightarrow}\ vwalk\ g\ {\isacharbrackleft}{\kern0pt}u{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbar}{\kern0pt}\isanewline
\ \ vwalk{\isacharunderscore}{\kern0pt}Cons{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}vwalk\ g\ {\isacharparenleft}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ has{\isacharunderscore}{\kern0pt}edge\ g\ u\ v\ {\isasymLongrightarrow}\ vwalk\ g\ {\isacharparenleft}{\kern0pt}u{\isacharhash}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline

### 2021-05-31 talk/terms/locale_primitives.tex
Finish slides

--- 2021-05-27 paper/terms/locale_primitives.tex
@@
 \ \ \ \ \isakeyword{and}\ vertex\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}v\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\isanewline
-\ \ \ \ \isakeyword{and}\ overlay\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\ {\isacharparenleft}{\kern0pt}\isakeyword{infixl}\ {\isacartoucheopen}{\isasymoplus}{\isacartoucheclose}\ {\isadigit{7}}{\isadigit{5}}{\isacharparenright}{\kern0pt}\isanewline
+\ \ \ \ \isakeyword{and}\ overlay\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\ {\isacharparenleft}{\kern0pt}\isakeyword{infixl}\ {\isacartoucheopen}{+}{\isacartoucheclose}\ {\isadigit{7}}{\isadigit{5}}{\isacharparenright}{\kern0pt}\isanewline
 \ \ \ \ \isakeyword{and}\ connect\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}g{\isachardoublequoteclose}\ {\isacharparenleft}{\kern0pt}\isakeyword{infixl}\ {\isacartoucheopen}{\isasymrightarrow}{\isacartoucheclose}\ {\isadigit{8}}{\isadigit{0}}{\isacharparenright}{\kern0pt}

### 2021-05-31 talk/terms/vwalk_def.tex
Finish slides

--- 2021-05-27 paper/terms/vwalk_def.tex
@@
 \ \ vwalk{\isacharunderscore}{\kern0pt}single{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}has{\isacharunderscore}{\kern0pt}vertex\ g\ u\ {\isasymLongrightarrow}\ vwalk\ g\ {\isacharbrackleft}{\kern0pt}u{\isacharbrackright}{\kern0pt}{\isachardoublequoteclose}\ {\isacharbar}{\kern0pt}\isanewline
-\ \ vwalk{\isacharunderscore}{\kern0pt}Cons{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}vwalk\ g\ {\isacharparenleft}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ has{\isacharunderscore}{\kern0pt}edge\ g\ u\ v\ {\isasymLongrightarrow}\ vwalk\ g\ {\isacharparenleft}{\kern0pt}u{\isacharhash}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
+\ \ vwalk{\isacharunderscore}{\kern0pt}Cons{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}vwalk\ g\ {\isacharparenleft}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ has{\isacharunderscore}{\kern0pt}edge\ g\ u\ v\ \isanewline{\isasymLongrightarrow}\ vwalk\ g\ {\isacharparenleft}{\kern0pt}u{\isacharhash}{\kern0pt}v{\isacharhash}{\kern0pt}xs{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline

### 2021-05-31 paper/paper.tex
Finish paper

@@
 adjacency lists employing immutable arrays~\cite{king1995dfs}. The
-consistency condition $E \subseteq V \times V$ is not checked statically though,
+\emph{consistency condition} $E \subseteq V \times V$ is not checked statically,
 which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
@@
 representation. We will also cover the algebraic structure (\ref{sec:algebra})
-exhibited by these graphs and elegant graph transformations (\ref{sec:trafo})
+exhibited by these primitives and elegant graph transformations (\ref{sec:trafo})
 based on a type class for the core. In~\autoref{sec:verification} we explore how
@@
 then are defined as follows. The \textit{empty} graph, denoted by $\eps$, is
-simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
+the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
 $\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
@@
 \end{align*}
-So overlay is just the union of two graphs, while connect also adds an edge from
+Overlay is just the union of two graphs, while connect also adds an edge from
 each vertex of its first argument to each vertex of its second argument. It is
@@
 graphs this already proves that algebraic graphs are a consistent
-representation for graphs, i.e.\ using the construction primitives only yields
+representation for graphs, i.e.\ the construction primitives only yield
 consistent graphs. The completeness of this representation, i.e.\ for
@@
 Before turning to more advanced graph constructions we will observe the
-algebraic structure of the primitives. It turns out that overlay and connect
-almost form a semiring over graphs:
+algebraic structure of the primitives. Overlay and connect form an algebra
+abiding the following axioms:
 \begin{itemize}
-\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
+\item $\oplus$ is commutative and associative
 \item $(\mathcal{G}, \connect, \eps)$ is a monoid
@@
 \end{itemize}
-For a proper semiring an annihilating zero for $\connect$ (i.e.\
-$x \connect 0 = 0 = 0 \connect x$) is missing. It is also unusual that the two operations
-share their identity $\eps$.
-
 In addition there is the following
@@
 With decomposition we can prove by equational reasoning that $\overlay$ is
-idempotent and that $\eps$ is indeed its identity.
-According to Mokhov, the remaining axioms are a minimal set of axioms
+idempotent and that $\eps$ is its identity. Thus, overlay and connect
+almost form a semiring over graphs:
+\begin{itemize}
+\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
+\item $(\mathcal{G}, \connect, \eps)$ is a monoid
+\item $\connect$ distributes over $\overlay$
+\end{itemize}
+For a proper semiring the annihilating zero for $\connect$ (i.e.\
+$x \connect \eps = \eps = \eps \connect x$) is missing. It is also unusual that the two operations
+share their identity $\eps$.
+
+According to Mokhov, the stated axioms are a minimal set of axioms
 \emph{characterizing} algebraic graphs~\cite{mokhov2017algebraic}. From the
@@
 $1\connect 1 = 1$ gives a structure abiding all axioms. This definition of
-$\mathcal{G}$ is clearly not (isomorphic to) what we call directed graphs for
-any universe of vertices.
+$\mathcal{G}$ corresponds to graphs with an empty universe of vertices (as there
+is only the empty graph). However, this definition is independent of the
+underlying universe of vertices. So we can define the same structure
+for a non-empty universe. In this scenario the primitives don't behave like what we
+call (isomorphic to) graphs.
 We don't know which additional axioms are required to obtain this isomorphism.
@@
 follow from the given axioms. Hence, adding an axiom scheme $\forall u \in
-\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ probably
-needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$), however, we don't know if this
-is sufficient.
+\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ 
+needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$).
 
@@
 \end{minted}
-They are a direct translation of the fact, that any algebraic graph expression
+They are a direct translation of the fact that any algebraic graph expression
 $g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
@@
 \begin{itemize}
-\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
+\item $\oplus$ is commutative and associative
 \item $\leftrightarrow$ is commutative and has $\eps$ as the identity
@@
 Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
-of \hs{+} and \hs{*} as shortcuts for \hs{overlay} resp.\ \hs{connect}.
+of \hs{+} and \hs{*} as shortcuts for \hs{overlay} and \hs{connect}.
 
@@
 \end{minted}
-Although for certain representations it is possible to get an $\mathcal{O}(1)$
-transpose operation (e.g.\ by setting a transposed flag), for many typical
-representations the whole graph has to be traversed, resulting in
-$\mathcal{O}(\abs{V} + \abs{E})$ time.
+$\mathcal{O}(1)$ transposition can be achieved by representations which allow to
+use a flag to treat edges as transposed. For many typical representations (e.g.\
+adjacency lists) the whole graph has to be traversed during transposition,
+resulting in an $\mathcal{O}(\abs{V} + \abs{E})$ time operation.
 
@@
 \hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
-\hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
+of type \texttt{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
 vertices of type \hs{b}.
@@
 \end{minted}
-Interestingly this can be used to merge vertices fulfilling some predicate
+Interestingly this can be used to contract vertices fulfilling some predicate
 \hs{p} by mapping them to a single vertex:
@@
   \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
-the definition of several \textit{locale}s. For the purposes of this paper a
-locale be thought of as a type class which can be equipped with axioms. The
+includes several \textit{locale}s. For the purposes of this paper a
+locale can be thought of as a type class which can be equipped with axioms. The
 following locale is the basis for our formalization.
@@
 challenge posed by the approach is that a graph is simply a value of some
-abstract type \Term{'g}. Thus, we characterize the existence of a vertex or an
-edge in a given graph by the subgraph relation (denoted by $\sqsubseteq$ in the
-formalization):
+type \Term{'g}. Thus, we characterize the existence of a vertex or an
+edge in a given graph by the subgraph relation.
+A graph $g$ contains a vertex $u$ if and only if $u \subseteq g$ and an edge
+$(u,v)$ if and only if $u \connect v \subseteq g$. The Isabelle definitions for
+this look as follows:
 
@@
 \vspace{3mm}\newline
+Where $\sqsubseteq$ is used instead of $\subseteq$ and \Term{edge u v} is
+defined as \Term{connect (vertex u) (vertex v)}.
 Equipped with these definitions we can then go on to inductively define when a
@@
 Isabelle) is a walk and $(u,v) \in E$, then prepending \Term{u} also yields a
-walk. Now we can prove various statements about walks (e.g.\ appending walks,
-splitting walks, etc.). Interestingly, many of these statements follow directly
+walk. Now we can prove various statements about walks (e.g.\ under which
+conditions walks can be appended, how walks can be split, etc.). Interestingly,
+many of these statements follow directly
 from the inductive definition of walks, i.e.\ we don't require any of the
@@
 Reachability then can be defined via the reflexive, transitive closure of the
-edge set. The second approach is to say that $v$ is reachable from $u$ if and
+edge set, while also considering every vertex can reach itself. The second approach is to say that $v$ is reachable from $u$ if and
 only if there is a walk from $u$ to $v$. These are, however, only proof of
@@
 graphs. We did not experiment with adding additional axioms as described
-in~\autoref{sec:algebra}. We instantiate the locale with the unit type, where
-all primitives simply map to the unit value. This formalizes that the current
-set of axioms is insufficient.
+in~\autoref{sec:algebra}. We do give an instantiation of the locale with the
+unit type, illustrating that the current axioms are not sufficient.
 
@@
 structure of graphs for polymorphic graph constructions and transformations, and
-for formal verification. However, the deep embedding of the
-construction primitives in a datatype is interesting in itself. Recall the
-definition from the introduction:
+for formal verification. Deeply embedding the construction primitives in a
+datatype is interesting in itself. The definition of the datatype is
+straightforward:
 \begin{minted}{haskell}
@@
 the way for further formalization efforts on the deep embedding. However, this
-also requires more work to become truly interesting, i.e.\ if algorithms that
-work directly on the deep embedding or other advanced usages of it can be found.
+also requires more work to become truly interesting, i.e.\ the discovery of algorithms that
+work directly on the deep embedding or other advanced usages.
 
@@
 However, they are using an algebra of relations as opposed to the one here
-defined on graph construction primitives. Dolan~\cite{dolan2013fun} uses an
-algebra on adjacency matrices (specifically a semiring) to find paths in graphs.
+defined on graph construction primitives. Dolan~\cite{dolan2013fun} uses a
+semiring on adjacency matrices to find paths in graphs.
 Mokhov himself postulates that the algebraic approach allows to formulate graph
